Interpolants and Error Bounds for Modeling and Predicting Variability in Computer Systems

### 2020-07-28 3-Disseration/error.tex
Updated thesis document, sent draft to Dr. Watson.

This chapter presents the theoretical results bounding the error of
(piecewise) linear interpolation. The error analysis relies on linear
interpolation for three reasons: (1) second order results can be
obtained utilizing a Lipschitz constant on the gradient of a function,
rather than standard Lipschitz bounds; (2) the results directly apply
to Delaunay interpolation; and (3) multiple other interpolants in this
work compute predictions as convex combinations of observed function
values, which may allow for straight forward extensions of this error
bound.

\begin{lemma}
  \label{lemma:1}
  Let $S \subset \mathbb{R}^d$ be open and convex, $f: \mathbb{R}^d
  \rightarrow \mathbb{R}$, and $\nabla f \in
  Lip_{(\gamma,\|\cdot\|_2)}(S)$, the set of $\gamma$-Lipschitz
  continuous functions in the $2$-norm. Then for all $x,y \in S$
  $$\big|f(y) - f(x) - \langle \nabla f(x), y - x \rangle \big| \leq \frac{\gamma \|y - x\|_2^2}{2}.$$
\end{lemma}

\begin{proof}
  Consider the function $g(t) = f \big((1-t) x + t y \big)$, $0 \leq t
  \leq 1$, whose derivative $g'(t) = \big\langle \nabla f \big((1-t) x
  + t y \big), y - x \big\rangle$ is the directional derivative of $f$
  in the direction $(y - x).$
  \begin{align*}
    \big|f(y) - f(x) - \langle \nabla f(x), y - x \rangle \big|
       &= \big|g(1) - g(0) - g'(0) \big| & \\
       &= \bigg| \int_0^1 g'(t) - g'(0)\ dt \bigg| \\
       &\leq \int_0^1 \big|g'(t) - g'(0)\big|\ dt \\
       &= \int_0^1 \bigg| \big \langle \nabla f\big((1-t)x + ty\big) - \nabla f(x), y - x \big \rangle \bigg|\ dt \\
       &\leq \int_0^1 \big \| \nabla f\big((1-t)x + ty\big) - \nabla f(x) \big \|_2\ \| y - x \|_2\ dt \\
       &\leq \int_0^1 \big ( \gamma\ \|y-x\|_2 \big) \ \big( \|y-x\|_2 \big) t\ dt \\
       &= \frac{\gamma \|y - x\|_2^2}{2}.
  \end{align*}
\end{proof}

\begin{lemma}
  \label{lemma:2}
  Let $x, y, v_i \in \mathbb{R}^d$, $c_i \in \mathbb{R}$, and
  $|\langle y - x, v_i \rangle| \leq c_i$ for $i = 1$, $\ldots$, $d.$
  If $M = (v_1$, $\ldots$, $v_d)$ is nonsingular, then
  $$\|y - x\|_2^2 \leq \frac{1}{\sigma_d^2} \sum_{i=1}^d c_i^2,$$
  where $\sigma_d$ is the smallest singular value of $M.$
\end{lemma}

\begin{proof}
  Using the facts that $M$ and $M^t$ have the same singular values,
  and $\|M^tw\|_2 \geq \sigma_d \|w\|_2$, gives
  \begin{align*}
    \|y - x\|_2^2 &\leq \frac{\|M^t (y - x)\|_2^2}{\sigma_d^2} \\
                  &=    \frac{1}{\sigma_d^2} \sum_{i=1}^d \langle y - x, v_i \rangle^2 \\
                  &\leq \frac{1}{\sigma_d^2} \sum_{i=1}^d c_i^2.
  \end{align*}
\end{proof}

\begin{lemma}
  \label{lemma:3}
  Given $f$, $\gamma$, $S$ as in {\it Lemma \ref{lemma:1}}, let $X =
  \{x_0$, $x_1$, $\ldots$, $x_d\}$ $\subset S$ be the vertices of a
  $d$-simplex, and let $\hat f(x) = \langle c, x - x_0 \rangle +
  f(x_0)$, $c \in \mathbb{R}^d$ be the linear function interpolating
  $f$ on $X.$ Let $\sigma_d$ be the smallest singular value of the
  matrix $M = (x_1 - x_0$, $\ldots$, $x_d - x_0)$, and $k =
  \max\limits_{1\ \leq\ j\ \leq\ d} \|x_j - x_0\|_2.$ Then
  $$\big\|\nabla f(x_0) - c\big\|_2 \leq \sqrt{d} \frac{\gamma k^2}{\sigma_d}.$$
\end{lemma}

\begin{proof}
  Consider $f(x) - \hat f(x)$ along the line segment $z(t) = (1-t)x_0
  + t x_j$, $0 \leq t \leq 1.$ By Rolle's Theorem, for some $0 < \hat
  t < 1$, $\big\langle \nabla f\big(z(\ \hat t\ )\big) - c, x_j - x_0
  \big\rangle = 0.$ Now
  \begin{align*}
    \big| \big< \nabla f(x_0) - c, x_j - x_0 \big \rangle \big|
        &= \big| \big< \nabla f(x_0) - \nabla f\big(z(\ \hat t\ )\big) + \nabla f\big(z(\ \hat t\ )\big) - c, x_j - x_0 \big \rangle \big| \\
	&= \big| \big\langle \nabla f(x_0) - \nabla f \big(z(\ \hat t\ )\big), x_j - x_0 \big \rangle \big| \\
	&\leq \big \| \nabla f(x_0) - \nabla f\big(z(\ \hat t\ )\big) \big \|_2 \|x_j - x_0\|_2 \\
	&\leq \gamma \|x_0 - z(\ \hat t\ )\|_2\ \|x_j - x_0\|_2 \\
	&\leq \gamma \|x_j - x_0\|_2^2 \leq \gamma k^2, \\
  \end{align*}
  for all $1 \leq j \leq d.$ Using {\it Lemma \ref{lemma:2}},
  $$\big \| \nabla f(x_i) - c \big \|_2^2 \leq \frac{d}{\sigma_d^2}
  \big( \gamma k^2\big)^2 \implies \big \| \nabla f(x_i) - c \big \|_2
  \leq \sqrt{d} \frac{\gamma k^2}{\sigma_d}.$$
\end{proof}

\begin{theorem}
  Under the assumptions of {\it Lemma \ref{lemma:1}} and {\it Lemma
    \ref{lemma:3}}, for $z \in S$,
  $$ \big|f(z) - \hat f(z)\big| \leq \frac{\gamma \|z - x_0\|_2^2}{2} + \sqrt{d} \frac{\gamma k^2}{\sigma_d} \|z - x_0\|_2.$$
\end{theorem}

\begin{proof}
  Let $v = \nabla f(x_0) - c$, where $\|v\|_2 \leq \sqrt{d} \gamma k^2 / \sigma_d$ by {\it Lemma \ref{lemma:3}}. Now
  \begin{align*}
    \big|f(z) - \hat f(z)\big|
       &= \big|f(z) - f(x_0) - \langle c, z - x_0 \rangle \big| \\ 
       &= \big|f(z) - f(x_0) - \langle \nabla f(x_0) - v, z - x_0 \rangle \big| \\
       &= \big|f(z) - f(x_0) - \langle \nabla f(x_0) , z - x_0 \rangle + \langle v , z - x_0 \rangle \big| \\
       &\leq \big|f(z) - f(x_0) - \langle \nabla f(x_0) , z - x_0 \rangle \big| + \big| \langle v , z - x_0 \rangle \big| \\
       &\leq \big|f(z) - f(x_0) - \langle \nabla f(x_0) , z - x_0 \rangle \big| + \|v\|_2 \|z - x_0\|_2 \\
       &\leq \big|f(z) - f(x_0) - \langle \nabla f(x_0), z - x_0 \rangle \big| + \textstyle{\frac{\gamma k^2 \sqrt{d}}{\sigma_d}} \|z - x_0\|_2 \\
       &\leq \frac{\gamma \|z - x_0\|_2^2}{2} + \sqrt{d}\frac{\gamma k^2}{\sigma_d} \|z - x_0\|_2,
  \end{align*}
  where the last inequality follows from {\it Lemma \ref{lemma:1}}.
\end{proof}

In summary, the approximation error of a linear (simplicial)
interpolant tends quadratically towards zero when approaching observed
data only when the diameter of the simplex is also reduced at a
proportional rate.  Only linear convergence to the true function can
be achieved in practice, without the incorporation of additional
observations. Notice that the approximation error is largely
determined by the spacing of observed data. Predictions made by
simplices whose vertices are not well-spaced (i.e., have large
diameter, or are nearly contained in a hyperplane) have higher
error. In light of this error bound, an empirical evaluation of
presented algorithms follows.

@@

The theoretical results constructed at the start of this chapter for
(piecewise) linear interpolation are promising and apply directly to
Delaunay interpolation, however they are difficult to interpret in
context with approximation algorithms that do not have similar known
uniform error bounds. For that reason, this chapter utilizes five
different data sets of varying dimension and application to construct
approximations and compare the accuracy of different techniques.

@@

In addition to the figures displaying approximation results for each
data set, tables of accompanying numerical results are located in the
Appendix. All of the test data sets capture underlying functions that
are almost certainly stochastic. As described in Section
\ref{sec:introduction}, regression techniques appear most appropriate
for these problems. However, typically data grows exponentially more
sparse with increasing dimension. Given sparse data, regressors tend
towards interpolation.

@@

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/NA/Fig2.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Forest Fire Area Burned};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
  \end{tikzpicture}
  \caption{Histogram of forest fire area burned under recorded weather
    conditions. The data is presented on a $\ln$ scale because
    most values are small with exponentially fewer fires on record
    that burn large areas.}
  \label{fig:hist-forest-fire}
\end{figure}

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.8\textwidth,]{Figures/NA/Fig3.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Area Burned Error};
  \end{tikzpicture}
  \caption{All models are applied to approximate the amount of area
    that would be burned given environment conditions. $10$-fold cross
    validation as described in the beginning of Section \ref{sec:error_data}
    is used to evaluated each algorithm. This results in exactly one
    prediction from each algorithm for each data point. These boxes
    depict the median (middle bar), median $95\%$ confidence interval
    (cones), quartiles (box edges), fences at $3/2$ interquartile
    range (whiskers), and outliers (dots) of absolute prediction error
    for each model. Similar to Figure \ref{fig:hist-forest-fire}, the
    errors are presented on a $\ln$ scale. The numerical data
    corresponding to this figure is provided in Table
    \ref{table:error-forest-fire} in the Appendix.}
  \label{fig:error-forest-fire}
\end{figure}

@@

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/NA/Fig4.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Parkinson's Total UPDRS Score};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
  \end{tikzpicture}
  \caption{Histogram of the Parkinson's patient total UPDRS clinical
    scores that will be approximated by each algorithm.}
  \label{fig:hist-parkinsons}
\end{figure}

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.8\textwidth,]{Figures/NA/Fig5.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Total UPDRS Score Error};
  \end{tikzpicture}
  \caption{All models are applied to approximate the total UPDRS score
    given audio features from patients' home life, using $10$-fold
    cross validation. These boxes depict the median (middle bar),
    median $95\%$ confidence interval (cones), quartiles (box edges),
    fences at $3/2$ interquartile range (whiskers), and outliers
    (dots) of absolute prediction error for each model. The numerical
    data corresponding to this figure is provided in Table
    \ref{table:error-parkinsons} in the Appendix.}
  \label{fig:error-parkinsons}
\end{figure}

@@

Delaunay achieves the lowest KS statistic for $62\%$ of
approximations, while Voronoi is best for the remaining $38\%$. Figure
\ref{fig:error-throughput} shows that while Delaunay may have more
best predictions, the behavior of Voronoi may be preferable.
%% -------------------------------------------------------------------

>>>>

This chapter presents theoretical results bounding the error of
(piecewise) linear interpolation. The error analysis relies on linear
interpolation for three reasons: (1) second order results can be
obtained utilizing a Lipschitz constant on the gradient of a function,
rather than standard Lipschitz bounds; (2) the results directly apply
to Delaunay interpolation; and (3) multiple other interpolants in this
paper compute predictions as convex combinations of observed function
values, which may allow for straightforward extensions of this error
bound.

\begin{plainlemma}
  \label{lemma:1}
  Let $S \subset \mathbb{R}^d$ be open and convex, $f: S \rightarrow
  \mathbb{R}$, and let $\nabla f(x)$ be $\gamma$-Lipschitz continuous
  in the $2$-norm. Then for all $x,y \in S$
  $$\big|f(y) - f(x) - \langle \nabla f(x), y - x \rangle \big| \leq \frac{\gamma \|y - x\|_2^2}{2}.$$
\end{plainlemma}

\begin{proofdot}
  Consider the function $g(t) = f \big((1-t) x + t y \big)$, $0 \leq t
  \leq 1$, whose derivative $g'(t) = \big\langle \nabla f \big((1-t) x
  + t y \big), y - x \big\rangle$ is the directional derivative of $f$
  in the direction $(y - x).$
  \begin{align*}
    \big|f(y) - f(x) - \langle &\nabla f(x), y - x \rangle \big|
        = \big|g(1) - g(0) - g'(0) \big| \\
       &= \bigg| \int_0^1 g'(t) - g'(0)\ dt \bigg| \leq \int_0^1 \big|g'(t) - g'(0)\big|\ dt \\
       &= \int_0^1 \bigg| \big \langle \nabla f\big((1-t)x + ty\big) - \nabla f(x), y - x \big \rangle \bigg|\ dt \\
       &\leq \int_0^1 \big \| \nabla f\big((1-t)x + ty\big) - \nabla f(x) \big \|_2\ \| y - x \|_2\ dt \\
       &\leq \int_0^1 \big ( \gamma\ \|y-x\|_2 \big) \ \big( \|y-x\|_2 \big) t\ dt = \frac{\gamma \|y - x\|_2^2}{2}.
  \end{align*}
  \qed
\end{proofdot}

\hfill

\begin{plainlemma}
  \label{lemma:2}
  Let $x, y, v_i \in \mathbb{R}^d$, $c_i \in \mathbb{R}$, and
  $|\langle y - x, v_i \rangle| \leq c_i$ for $i = 1$, $\ldots$, $d.$
  If $M = (v_1$, $\ldots$, $v_d)$ is nonsingular, then
  $$\|y - x\|_2^2 \leq \frac{1}{\sigma_d^2} \sum_{i=1}^d c_i^2,$$
  where $\sigma_d$ is the smallest singular value of $M.$
\end{plainlemma}

\begin{proofdot}
  Using the facts that $M$ and $M^t$ have the same singular values,
  and $\|M^tw\|_2 \geq \sigma_d \|w\|_2$, gives
  \begin{align*}
    \|y - x\|_2^2 &\leq \frac{\|M^t (y - x)\|_2^2}{\sigma_d^2} \\
                  &=    \frac{1}{\sigma_d^2} \sum_{i=1}^d \langle y - x, v_i \rangle^2 \\
                  &\leq \frac{1}{\sigma_d^2} \sum_{i=1}^d c_i^2.
  \end{align*}
  \qed
\end{proofdot}

%% \newpage

\begin{plainlemma}
  \label{lemma:3}
  Given $f$, $\gamma$, $S$ as in {\it Lemma \ref{lemma:1}}, let $X =
  \{x_0$, $x_1$, $\ldots$, $x_d\}$ $\subset S$ be the vertices of a
  $d$-simplex, and let $\hat f(x) = \langle c, x - x_0 \rangle +
  f(x_0)$, $c \in \mathbb{R}^d$ be the linear function interpolating
  $f$ on $X.$ Let $\sigma_d$ be the smallest singular value of the
  matrix $M = (x_1 - x_0$, $\ldots$, $x_d - x_0)$, and $k =
  \max\limits_{1\ \leq\ j\ \leq\ d} \|x_j - x_0\|_2.$ Then
  $$\big\|\nabla f(x_0) - c\big\|_2 \leq \frac{\sqrt{d} \, \gamma k^2}{2 \sigma_d}.$$
\end{plainlemma}

\begin{proofdot}
  \def\g{\gamma_g}
  Consider $g(t)=f(x(t)) - \hat f(x(t))$ along
  the line segment $x(t)=(1-t)x_0 + tx_j$, $0\le t\le1$, from $x_0$ to
  $x_j$.  Observe that $g(0)=f(x_0) - \hat f(x_0) = 0$, $g(1)=f(x_j) -
  \hat f(x_j) = 0$, and $g'(t)$ is $\g$-Lipschitz continuous with $\g
  = \gamma \|x_j-x_0\|_2^2$. The following is visualized in 
  Figure \ref{fig:lemma3}.

  Suppose $g'(0) > \g/2$. Then $|g'(0)-g'(t)| \le \g t \implies g'(0)
  - \g t \le g'(t)$, and the line $w=g'(0)-\g t$ intersects the
  $t$-axis at $\tilde t=g'(0)/\g > 1/2$. $\tilde t<1$ necessarily,
  since by Rolle's Theorem there exists $0<z<1$ such that $g'(z)=0$
  and so $g'(0) - \g z \le g'(z) = 0$. Now by integrating
  $$\frac{g'(0)^2}{2\g} = \frac{g'(0)\tilde t}{2} = \int_0^{\tilde t}
  (g'(0)-\g t)dt \le \int_0^{\tilde t} g'(t)dt = g(\tilde t),$$ and
  using $g'(0) > \g / 2$
  \begin{align*}
    \frac{-g'(0)^2}{2\g} < g'(0) - \frac{g'(0)^2}{2\g} - \g/2 =
    \frac{(1-\tilde t)(g'(0)-\g)}{2} &= \int_{\tilde t}^1 (g'(0)-\g
    t)dt \\ &\le \int_{\tilde t}^1 g'(t)dt = -g(\tilde t),
  \end{align*}
  \noindent a contradiction for the value of $g(\tilde t)$. A similar
  contradiction arises for $g'(0)<-\g/2$. Therefore $|g'(0)| \le
  \g/2$. In terms of $f$,
  $$\bigl|\langle \nabla f(x_0)-c,x_j-x_0\rangle\bigr| = |g'(0)| \le
  \gamma \|x_j-x_0\|_2^2/2 \le \gamma k^2 / 2,$$ which holds for all
  $1\le j\le d$.  Finally, using {\it Lemma \ref{lemma:2}},
  $$\|\nabla f(x_0)-c\|_2^2 \le \frac{d}{\sigma_d^2}(\gamma k^2/2)^2
  \implies \|\nabla f(x_0)-c\|_2 \le \frac{\sqrt{d} \, \gamma
    k^2}{2\sigma_d}.$$ \qed
\end{proofdot}

\begin{figure}
  \centering
  \includegraphics[width=0.9\textwidth,]{Figures/NA/example_lemma_3.pdf}
  \def\g{\gamma_g}
  \caption{Three different scenarios visualizing {\it
      Lemma \ref{lemma:3}}, where $g(t)$ is the difference between a
    piecewise linear interpolant and the approximated function along a
    normalized line segment between interpolation points, $g'(t)$ is
    $\g$-Lipschitz continuous, and $w$ and $\tilde t$ are defined in
    the proof. Leftmost is a randomly chosen permissible shape of $g$
    and $g'$. The middle is the only possible shape of $g$ and $g'$
    given $g'(0) = \g / 2$, establishing the case of equality in the
    lemma. Rightmost is the resulting contradiction when $g'(0) > \g /
    2$, notice it is impossible to ensure $g'(t)$ is $\g$-Lipschitz
    continuous and satisfy $g(1) = 0$ (highlighted with red circle on
    the right).
  \vspace{-.1cm}}
  \label{fig:lemma3}
\end{figure}

\begin{onetheorem}
  Under the assumptions of {\it Lemma \ref{lemma:1}} and {\it Lemma
    \ref{lemma:3}}, for $z \in S$,
  $$ \big|f(z) - \hat f(z)\big| \leq \frac{\gamma \|z - x_0\|_2^2}{2} + \frac{\sqrt{d} \, \gamma k^2}{2 \sigma_d} \|z - x_0\|_2.$$
\end{onetheorem}

\begin{proofdot}
  Let $v = \nabla f(x_0) - c$, where $\|v\|_2 \leq \sqrt{d}\, \gamma k^2 / (2 \sigma_d)$ by {\it Lemma \ref{lemma:3}}. Now
  \begin{align*}
    \big|f(z) - \hat f(z)\big|
       &= \big|f(z) - f(x_0) - \langle c, z - x_0 \rangle \big| \\ 
       &= \big|f(z) - f(x_0) - \langle \nabla f(x_0) - v, z - x_0 \rangle \big| \\
       &= \big|f(z) - f(x_0) - \langle \nabla f(x_0) , z - x_0 \rangle + \langle v , z - x_0 \rangle \big| \\
       &\leq \big|f(z) - f(x_0) - \langle \nabla f(x_0) , z - x_0 \rangle \big| + \big| \langle v , z - x_0 \rangle \big| \\
       &\leq \big|f(z) - f(x_0) - \langle \nabla f(x_0) , z - x_0 \rangle \big| + \|v\|_2 \|z - x_0\|_2 \\
       &\leq \big|f(z) - f(x_0) - \langle \nabla f(x_0), z - x_0 \rangle \big| + \big(\sqrt{d}\,\gamma k^2 / (2 \sigma_d)\big) \|z - x_0\|_2 \\
       &\leq \frac{\gamma \|z - x_0\|_2^2}{2} + \frac{\sqrt{d}\,\gamma k^2}{2 \sigma_d} \|z - x_0\|_2,
  \end{align*}
  where the last inequality follows from {\it Lemma \ref{lemma:1}}.
  %
  \qed
\end{proofdot}

In summary, the approximation error of a linear (simplicial)
interpolant tends quadratically towards zero when approaching observed
data only when the diameter of the simplex is also reduced at a
proportional rate. Only linear convergence to the true function
is achieved in practice without the
incorporation of additional observations, by moving closer to
  interpolation points. Notice that the approximation error is
largely determined by the spacing of observed data. Predictions made
by simplices whose vertices are not well-spaced (i.e., have large
diameter, or are nearly contained in a hyperplane) have higher error.

That the theoretical error bound is sharp can be observed with the
test function $q(x) = \|x\|_2^2,$ $x \in \mathbb{R}^d,$ and the
simplex defined by vertices $X = \{0, e_1, \ldots, e_d\},$ where $e_i$
is the $i$-th standard basis vector in $\mathbb{R}^d,$ $e = (1$,
$\ldots$, $1) \in \mathbb{R}^d,$ and $0$ denotes the zero vector in
any dimension.  The constants relevant to the error bound are
$$ \gamma = 2, \qquad \sigma_d = 1, \qquad k = 1, \qquad x_0 = 0,
\qquad \hat q(x) = \langle e, x - 0\rangle + q(0). $$
\noindent Noting that $q(0) = 0,$ the approximation error at $z =
-(1/2)e$ is
$$ \big|q(z) - \hat q(z)\big| = \big|\|z\|_2^2 - \langle e,z
\rangle\big| = \big| d/4 + d/2 \big| = 3d/4, $$
\noindent while the error bound from the theorem gives

\vspace{-2mm}
\begin{align*}
  \big|q(z) - \hat q(z)\big|
  &\leq \frac{\gamma \|z - x_0\|_2^2}{2} + \frac{\sqrt{d}\,\gamma k}
      {2 \sigma_d} \|z - x_0\|_2 \\
  &{}=  \|z\|_2^2 + \sqrt{d} \|z\|_2 \\
  &{}=  d/4 + d/2 = 3d/4.
\end{align*}
\vspace{-2mm}

\noindent Acknowledging that the error bound is sharp, it may be of
interest to observe the error of piecewise linear approximation
techniques on an analytic test function.

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.9\textwidth,]{Figures/NA/Fig2.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Number of Data Points};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Absolute Error};
  \end{tikzpicture}
  \caption{Delaunay and MLP approximations are constructed from Fekete
    points over the unit cube evaluating the test function $f(x) =
    \cos(\|x\|_2)$ for $x \in \mathbb{R}^2$. The figure shows the
    first/third quartiles at the box bottom/top, the second quartile
    (median) at the white bar, median 95\% confidence interval (cones,
    barely visible in figure), and whiskers at 3/2 of the adjacent
    interquartile ranges, for the absolute prediction error for each
    model at $1000$ random evaluation points. The left plot observes a
    perfect interpolation problem with exact evaluations of $f.$ The
    right plot observes a regression problem with uniform random noise
    giving values in $[.9 f(x),\ 1.1f(x)]$ for each $x.$ Both axes are
    log scaled.}
  \label{fig:convergence-2d}
\end{figure}

%% ===================================================================
\vspace{-2mm}
\section{Demonstration on an Analytic Test Function}
\label{sec:analytic}

The theoretical results constructed in Section \ref{sec:theory} for
(piecewise) linear interpolation are promising and apply directly to
Delaunay interpolation, however they are difficult to interpret in
context with approximation algorithms that do not have similar known
uniform error bounds. For that reason, an analytic function is used to
measure the error of a piecewise linear interpolant (Delaunay) and a
piecewise linear regressor (the MLP) when provided an increasing
amount of data with varying levels of random noise. Only
Delaunay and the MLP are considered in this demonstration because
all other mentioned techniques are not strictly piecewise linear
approximations.

The test function chosen here for analysis resembles the
\textit{oscillatory} function used by \cite{barthelmann2000high}.
However, a slight modification is made to remove the simple dot
product structure (which is favorable for the MLP). Let $f(x)
=\cos(\|x\|_2)$ for $x \in \mathbb{R}^d$ where $d$ is either $2$
(Fig. \ref{fig:convergence-2d}) or $20$
(Fig. \ref{fig:convergence-20d}). This function has a bounded change
in gradient $2$-norm and hence meets the necessary Lipschitz condition
for the error bound. Data points and approximation points for this
test will be within the unit cube $[0,1]^d$.

The error of a linear interpolant constructed from well spaced points
is dominated by the distance to the nearest data point. In order to
uniformly decrease the distance to the nearest data point across the
unit cube, exponentially more data points must be available for
approximation. For this experiment $N = 2(4^i),$ $i = 1,$ $\ldots,$
$6,$ chosen points are kept well spaced by computing approximate
Fekete points. The following experiment was also run with a
  Latin hypercube design \cite{park1994optimal}, which produced almost
  identical results that are not reported here. Fekete points have
a history in potential theory \cite{kovari_pommerenke_1968} and are
most generally defined as those points that maximize the absolute
value of a Vandermonde determinant. Here, the QR method outlined in
\cite{bos2010computing} is used to identify approximate Fekete points
from a Vandermonde matrix. A multivariate polynomial basis of size
$2N$ is constructed containing all polynomials of degree $\leq n$ for
the largest $n$ such that ${d+n \choose n} \leq 2N$ and $2N - {d+n
  \choose n}$ arbitrarily selected polynomials of degree $n + 1.$ The
Vandermonde matrix for these $2N$ polynomials is used to select $N$
approximate Fekete points.

Finally an additional aspect is added to this test problem by
incorporating random uniform noise into the evaluations of $f.$ For
each test two experiments are executed, one with exact function
evaluations (an interpolation problem) and one with a
constant signal-to-noise ratio (SNR) of 10:1 (a regression
  problem).

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.9\textwidth,]{Figures/NA/Fig3.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Number of Data Points};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Absolute Error};
  \end{tikzpicture}
  \caption{Delaunay and MLP approximations are constructed from Fekete
    points over the unit cube evaluating the test function $f(x) =
    \cos(\|x\|_2)$ for $x \in \mathbb{R}^{20}$. The details are the
    same as for Fig. \ref{fig:convergence-2d}.}
  \label{fig:convergence-20d}
\end{figure}

The bound from the theorem suggests that by increasing the number of
well-spaced data points, approximation error can be reliably
decreased. The $d = 2$ test seen on the left half of Figure
\ref{fig:convergence-2d} shows a consistent decrease in error for
Delaunay and also shows the eventual accuracy plateau obtained by a
parametric regression form (the MLP, at roughly 500 points). On the
right hand side of Figure \ref{fig:convergence-2d} the random noise
clearly prohibits Delaunay from converging to $f$, while the MLP is
able to improve its approximation with more data points on average.
The convergence result for a very low-dimensional problem like this is
expected.  However, intuition fails for higher dimensional problems.

Figure \ref{fig:convergence-20d} shows the test function with $d = 20$
presents a significantly more challenging approximation problem than
its counterpart in low dimension. The same increase in number of data
points from 32 to 8192 causes no apparent improvement in approximation
for either the noise-free or the noisy problems. Perhaps unexpectedly,
the interpolation technique performs better than the regression
technique on the noisy data (right) in Figure
\ref{fig:convergence-20d}, and worse than the regression technique on
the noise-free data (left). This result emphasizes the relevance of
interpolation for problems in high dimension. It also reveals that the
outcome of MLP regressions can get worse when adding more data points.

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.9\textwidth,]{Figures/NA/theorem-convergence.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Number of Data Points};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Error, Distance, and Smallest SV };
  \end{tikzpicture}
  \caption{The distribution of absolute error, distance to
    the nearest data point, smallest singular value (SV) and the
    longest edge of the simplex containing each approximation point in
    the tests from Fig. \ref{fig:convergence-2d} and
    Fig. \ref{fig:convergence-20d} for Delaunay. In two dimensions it
    can be seen that $\|z - x_0\|_2$, $\sigma_d$, and $k$ all shrink
    at the same rate for well-spaced approximation points, resulting
    in a faster rate of decrease for approximation error. Notice that
    in higher dimension the data remains sparse even with thousands of
    data points, and the decay in data spacing is more prominant. The
    relatively small reduction in $k$ along with the decrease in
    $\sigma_d$ explain the minimal reduction in error seen by Delaunay
    in Fig. \ref{fig:convergence-20d}.}
  \label{fig:data-spacing}
\end{figure}

The results achieved for both $d = 2$ and $d = 20$ align with the
theoretical error bound as can be seen in Figure
\ref{fig:data-spacing}. In low dimension, thousands of data points
meaningfully reduce the sparsity of the approximation problem.
However, in higher dimension it takes many more points to achieve a
reduction in data sparsity while simultaneously being more difficult
to produce well-conditioned simplices. This evidences the
inherent challenge of data sparsity in high dimension approximation
problems. The analytic results presented are not caused by the chosen
test function, but rather the exponential increase in complexity that
accompanies increased dimension.

In summary, the regime of signal-to-noise ratios that result in
competitively accurate interpolants is greater for moderate to high
dimensional problems due to increased sparsity. Acknowledging the
viability of interpolation for problems of moderate dimension, the
next section will consider real-world problems of similar proportion
(thousands of examples in tens of dimensions).

@@

This section extends the comparison of interpolation and regression
algorithms to a sample of real-world problems. Five different data
sets of varying dimension and application are utilized to construct
approximations and compare the accuracy of different techniques.

@@

In addition to the figures displaying approximation results for each
data set, scatter plots of predicted versus actual values
  and tables of accompanying numerical results including
  timings and quartiles of prediction error are located in the
Appendix (Section \ref{sec:appendix}). All of the test data sets
capture underlying functions that are almost certainly stochastic. As
described in Section \ref{sec:introduction}, regression techniques
appear most appropriate for these problems. However, typically data
grows exponentially more sparse with increasing dimension. Given that
sparse data regressors tend towards interpolation and, as demonstrated
in Section \ref{sec:analytic}, interpolants produce similar (if not
identical) results, it is presumed that interpolants are equally
viable approximation techniques on these problems.

@@

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/NA/Fig4.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Forest Fire Area Burned};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
  \end{tikzpicture}
  \caption{Histogram of forest fire area burned under recorded weather
    conditions. The data is presented on a $\ln$ scale because
    most values are small with exponentially fewer fires on record
    that burn large areas.}
  \label{fig:hist-forest-fire}

  \vspace{.3cm}

  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.8\textwidth,]{Figures/NA/Fig5.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Area Burned Error};
  \end{tikzpicture}
  \caption{All models are applied to approximate the amount of area
    that would be burned given environment conditions. $10$-fold cross
    validation as described in the beginning of Section \ref{sec:error_data}
    is used to evaluated each algorithm. This results in exactly one
    prediction from each algorithm for each data point. These boxes
    depict the median (middle bar), median $95\%$ confidence interval
    (cones), quartiles (box edges), fences at $3/2$ interquartile
    range (whiskers), and outliers (dots) of absolute prediction error
    for each model. Similar to Figure \ref{fig:hist-forest-fire}, the
    errors are presented on a $\ln$ scale. The numerical data
    corresponding to this figure is provided in Table
    \ref{table:error-forest-fire} in the Appendix.}
  \label{fig:error-forest-fire}
\end{figure}

@@

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.75\textwidth,]{Figures/NA/Fig6.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Parkinson's Total UPDRS Score};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
  \end{tikzpicture}
  \caption{Histogram of the Parkinson's patient total UPDRS clinical
    scores that will be approximated by each algorithm.}
  \label{fig:hist-parkinsons}

  \vspace{.3cm}

  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.8\textwidth,]{Figures/NA/Fig7.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Total UPDRS Score Error};
  \end{tikzpicture}
  \caption{All models are applied to approximate the total UPDRS score
    given audio features from patients' home life, using $10$-fold
    cross validation. These boxes depict the median (middle bar),
    median $95\%$ confidence interval (cones), quartiles (box edges),
    fences at $3/2$ interquartile range (whiskers), and outliers
    (dots) of absolute prediction error for each model. The numerical
    data corresponding to this figure is provided in Table
    \ref{table:error-parkinsons} in the Appendix.}
  \label{fig:error-parkinsons}
\end{figure}

@@

Delaunay achieves the lowest KS statistic (max norm difference) for
$62\%$ of approximations, while Voronoi is best for the remaining
$38\%$. Figure \ref{fig:error-throughput} shows that while Delaunay
may have more best predictions, the behavior of Voronoi may be
preferable.
%% -------------------------------------------------------------------
